---
title: 'PEPR: Pipeline for Evaluating Prokaryotic References'
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
  html_document:
    fig_caption: yes
    keep_md: yes
    number_sections: yes
  word_document: default
bibliography: pepr-pub.bib
---

```{r, echo = FALSE, message=FALSE}
library(png)
library(grid)
library(peprr)
library(dplyr)
library(knitr)
source("rm_metadata.R")
source("calc_results_values.R")
```

```{r, echo = FALSE, message=FALSE}
incCount <- function(inObj, useName) {
    nObj <- length(inObj)
    useNum <- max(inObj) + 1
    inObj <- c(inObj, useNum)
    names(inObj)[nObj + 1] <- useName
    inObj
}
figCount <- c(`_` = 0)
tableCount <- c(`_` = 0)

pasteLabel <- function(preText, inObj, objName, insLink = TRUE) {
    objNum <- inObj[objName]

    useText <- paste(preText, objNum, sep = " ")
    if (insLink) {
        useText <- paste("[", useText, "](#", objName, ")", sep = "")
    }
    useText
}
```


# Authors
Nathan D. Olson
Justin Zook
Scott A. Jackson
Marc L. Salit

# Abstract

# Introduction
* General background - painting the picture
    * Advances in genome sequencing 
    * Genome sequencing applications
* Genome sequencing as a measurement process
    * for each step briefly describe known sources of bias and error
    * sample preparation
    * reaction monitoring
    * computational analysis
    * Complex measurement process requiring controls
    * New maturing field, unknown sources of bias and error.  Use of homogeneous and stable reference materials can be used in benchmarking.  Help to identify and characterize currently unknown sources of bias and error.
    
* Whole genome sequencing controls
    * Needed to validate measurement process
    * Microbial sequencing control material for regular QC, small scale relative to human genome material 
    * NIST developing RMs to meet this need
    * Brief description of micro RMs and approach to characterization
        * RM characterization 
            * homogeneity, stability, purity
            * genome assembly
            * base level analysis
            * genomic purity
        * Characterization philosophy
            * maturing field - basic research still needed to be able to quantify and define confidence in a genome sequence
            * use of replicates and orthogonal measurements
            * providing genome sequence and characterization results allow users to investigate questionable regions and positions positions 
    
* Also need for application specific materials
    * Sequencing manufacturers - material with characteristics that best challenge their platforms limitations
    * Application areas - validation of methods for detection and characterization of specific pathogens
    * Different genome contexts
    * Limited resources to create RMs, 
        * user characterized materials
        * help to identify other sources of bias and error for genome context not found in NIST RMs
    * Sequencing centers and platform manufactures already using material for QC
    

    NIST RM characterization pipeline was developed so that users can characterize material they are currently using as part of a mature quality control system.  Application of the pipeline increases the confidence in the quality control system through better characterization of the control material.  Here we present a Pipeline for Evaluating Prokaryotic References "PEPR" and preliminary results from the characterization of the first of four microbial reference materials using the pipeline.

# Methods
## Pipeline for Evaluating Prokaryotic References: PEPR
Pipeline for Evaluating Prokaryotic Reference (PEPR) consists of three bioinformatic pipelines written in python `r I(pasteLabel("Figure", figCount, "f.pepr_workflow"))`.  The three bioinformatic pipelines are Genome Evaluation, Genome Characterization, and Genomic Purity.  Bioinformatic pipeline inputs defined with a yaml file.  The individual pipeline coordinates the execution of a number of commandline tools, ogging the standard output and standard error for each step in the pipeline  in time stamped files for later inspection and debugging. Pipeline code avaiable at ([github.com/nate-d-olson/pepr](https://github.com/nate-d-olson/pepr) ). To reduce the barrier for reuse two Docker ([www.docker.com](https://www.docker.com/) ) containers are available with pipeline dependencies pre-installed excluding the Genome Analysis Toolkit (due to licensing restrictions). Docker is a lightweight virtual environment that facilitates the sharing and distribution of computing environments that can be run on any desktop, cloud, and high performance computing environment, regardless of the operating system. The pepr container (https://registry.hub.docker.com/u/natedolson/pepr/) includes dependencies for the genome evaluation and characterization pipelines, a Dockerfile for building the container is included in the pepr repository. The docker-pathoscope ([https://registry.hub.docker.com/u/natedolson/docker-pathoscope/](https://registry.hub.docker.com/u/natedolson/docker-pathoscope/) ) has dependencies for the genomic purity pipeline installed.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.cap= "__PEPR workflow.__ White objects are pipeline inputs, grey objects are the three pipeline components, light blue objects are the pipeline products."}
figCounts <- incCount(figCount, "f.pepr_workflow")
grid.raster(readPNG("pepr-workflow.png"))
```

A software package peprr ([github.com/nate-d-olson/peprr](github https://github.com/nate-d-olson/peprr) ) was developed for the statistical computing languge R to compile the output from the genome evaluation, characterization, and genomic purity pipelines [@R2015]. The compiled data was manipulated into a series of data tables within an sqlite database to facilitate analysis [@wickham2014tidy].  Functions to generate a number of summary tables and figures, including those in this publication are included in the package.

## Pipeline Specific Methods
### Genome Evaluation Pipeline
The Genome Evaluation pipeline is the first step in the PEPR workflow and uses whole genome sequencing data to evaluate and correct misasseblies in the reference genome. Sequencing data are first retrieved from the Genbank Sequence Read Archive (SRA) using the sratoolkit fastq-dump ([http://ncbi.github.io/sra-tools/](http://ncbi.github.io/sra-tools/) ). The sequencing reads are then mapped to the reference genome using bwa (MiSeq and PacBio)[@Li2013c] and tmap [@Homer]. Finally Pilon was used to evaluate and polish the reference assembly assembly [@Walker2014].  The revised reference genome is then further characterized in the Genome Characterization Pipeline.

### Genome Characterization Pipeline
The Genome Characterication Pipeline uses replicate sequence datasets from multiple sequencing platforms for evaluate the reference genome on a individual base level. Seqeuncing data is aligned to the reference genome using the same methods as the Genome Evaluation pipeline. Next the sequence alignment files are processed prior to downstream analysis by marking duplicates with Picard's MarkDuplicates ([http://broadinstitute.github.io/picard](http://broadinstitute.github.io/picard) ), realignment around indels with the GenomeAnalysisToolKit [@McKenna2010; @DePristo12011]. The base level summary statistics are calculated using SAMtools mpileup [@Li2009].  Homogeneity analysis, measure of the similarity of the genomic content between replicate samples, was performed by first generating a pileup file for each dataset then performing pairwise tumor-normal variant calling using Varscan [@Koboldt2009]. Additionally, a number of summary statistics are caclulated for the sequencing datasets and alignment files, using fastqc ([http://www.bioinformatics.babraham.ac.uk/projects/fastqc/](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/) ) and Picard Collect Multiple Metrics ([http://broadinstitute.github.io/picard](http://broadinstitute.github.io/picard) ) respectively.  

### Genomic Purity Pipeline
The purity of the genomic material in terms of presense of DNA from sources other than the expected strain was assessed using the metagenomic taxonomic read classification algorithm PathoScope 2.0 [@Hong2014].  This method uses an expectation maximization algorithm where the sequence data are first mapped to a database comprised on all sequence data in the Genbank nt database, then through an iterative process re-assigns ambiguously mapped reads to based on the proportion of reads mapped unambigously to individual taxa in the database. The PathoScope 2.0 taxonomic read classification pipeline includes an initial read filtering step (PathoQC), followed by mapping reads to a reference database (PathoMap - a wrapper for bowtie2 [@Langmead2012]), then an expectation-maximization classification algorithm (PathoID).  The annotated Genbank nt database provided by the PathoScope developers was used as the reference database ([ftp://pathoscope.bumc.bu.edu/data/nt_ti.fa.gz](ftp://pathoscope.bumc.bu.edu/data/nt_ti.fa.gz)). 
    
## _S. aureus_ Sequencing
### Material
RM8376 is extracted genomic DNA from _Staphylococcus areus_ strain NRS100 isolate COL (Biosample SAMN02854573).  The strain was selected based on public health relevance and it's low \%GC.  The strain was selected for use by the FDA and isolated from a clinical sample by Children's National Hospital (Biosample SAMN02700075).  The strain is resistant to the antibiotics oxacillin and tetracycline.

A single colony was obtained from the initial culture stab after an overnight incubation at 37°C on an plate. A single colony was used to innoculate a new plate. One colony from the new plate was grown in 20 mL LB at 37°C. The culture was used to inoculate 15 X 150 mm plates which were incubated at 37°C for 16 hours. Glycerol freezes were made from the remaining stock. DNA and agarose plugs were made from the 15 x 150 mm plates. DNA was isolated by lysing the bacteria in lysis Solution containing NaCl, Tris, EDTA and lysostaphin (25 µ g/ml) and SDS. Proteinase K and RNase A were used to treat protein and RNA. Ammonium acetate was used to remove protein. DNA was recovered by isopropanol precipitation. DNA was washed with 70\% alcohol. Drained, then dissolved in TE (Tris 10 mM, EDTA 0.1 mM, pH8.0).

### Experimental Design
The _S. aureus_ candidate reference materials were sequenced using three orthogonal sequences methods. Long read sequencing using PacBio RSII[^pacbio], DNA Template Prep Kit 3.0 was used to prepare the library and sequenced with the P6-C4 chemistry.  Eight replicate vials were randomly sampled from the lot of 1500 vials and sequenced using two short read sequencing technologies, Ion Torrent PGM [^iontorrent], and Illumina MiSeq [^illumina]. Technical replicate sequencing libraries were prepared for each of the eight vials using the Nextera DNA Sample Prep Kit[^illumina], samples were barcoded using the Nextera Index Kit and sequenced using the MiSeq 600 cycle Reagent kit v3 for 2 X 300 bp reads.  The 16 libraries were pooled and sequencing in a single run.  Single barcoded 400 bp Ion Torrent PGM libraries were prepared for each of the eight vials using the Ion Xpress Plus kit [^iontorrent], including the Ion Xpress plus fragment plus library kit, with a 318C chip. The vials were barcoded using the IonXpress Kit[^iontorrent] and sequencing template was prepared using the Ion PGM Template OT2 400 kit and the IonPGM400 kit was used for sequencing. The raw sequence data is archived at the Genbank Sequence Read Archive ([ncbi.nlm.nih.gov/sra](http://www.ncbi.nlm.nih.gov/sra)), see `r I(pasteLabel("Table", tableCount, "t.seq_summary"))` for accession numbers. 

## Reference assembly
PacBio long read data was used to generate the reference genome assembly used as input with the Genome Evaluation Pipeline. PacBio Reads wereassembled using HGAP assembler - version __TODO__ [@Koren2013].  The _de novo_ assembly was validated using OpGen optical mapping data generate [^opgen].  Agarose plugs generated from the same culture stock as that used to generate the DNA reference material was used for optical mapping measurement.  Restriction enzyme NcoI was used for the restriction digest.

# Results
```{r echo=FALSE, message=FALSE} 
peprDB <- dplyr::src_sqlite(db_path)
```
## Summary of Sequencing Datasets
The number of reads, read, length, insert size for paired-end datasets as well as coverage for _S. aureus_ datasets, presented in Table 1, was generated from the database using the `seq_summary_table(peprDB)` command in the peprr package `r I(pasteLabel("Table", tableCount, "t.seq_summary"))`.  The MiSeq sequencing run resulted in over 1.5 million pair end (3 million total) reads per library with a median read length just under 250 bp.  Whereas the PGM sequencing run produced about 0.5 million reads per library with a median read length around 235 bp.  The higher throughput and paried end reads resulted in a higher per library coverage for MiSeq compared to PGM (~140X vs. ~30X).  Despite using the 400bp sequencing chemistry the PGM median read length was less than 250 bp.  The three PacBio datasets are technical sequencing replicates (SMRT cells) from the same sequencing library, with a median read lengths of XYZ, and throughput of XYZ, resulting in a total of ??X coverage.

```{r echo = FALSE, message=FALSE, warning=FALSE}
tableCount <- incCount(tableCount, "t.seq_summary")
kable(seq_summary_table(peprDB), 
      round = 0, row.names = FALSE, caption = "Summary of sequencing datasets")
```

## Genome Assembly
The _de-novo_ assembly using PacBio data was performed using the HGAP assembler, as part of the SMRTanalysis software (__version info__), resulting in a closed genome and plasmid assembly. Optical mapping data generated by the OpGen Argus system was used to validate the genome assembly.  Optical mapping data agreed with the _de novo_ assembly `r I(pasteLabel("Figure", figCount, "f.assembly_comp"))`. The plasmid is too small for OpGen optical mapping technology and therefore the plasmid assmebly was not validated with the optical mapping data.  

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.scap= "Comparison of optical map data to genome assembly.", fig.cap = "Alignment of in-silico genome map generated from the PacBio HGAP assembly to the OpGen optical map.  Blue bars in map represent NocI restriction sites, black lines indicate co-linear regions."}
figCounts <- incCount(figCount, "f.assembly_comp")
grid.raster(readPNG("opgen_assembly_comparison.png"))
```

The short read sequencing data supported the _de novo_ genome assembly. Pilon, a program developed for genome finishing and variant calling, identifies positions in the a genome assembly that exbiting characteristics of missassemblies and base call errors.  Running Pilon using the MiSeq data did not identify any missassemblies or base call errors.  Pilon did identifed __TODO ## TODO__ one and two base indels, and __TODO ## TODO__ larger deletions in the reference assembly.  Evaluation of the mapped MiSeq and PGM reads to these regions identicates that __TODO figures and summary statement TODO___.

## Base Level Analysis
### Base Purity
A base purity metric was used for base level analysis.  The purity value is the number bases in reads aligned to a genome position that are in agreement with the reference base divided by the total number of reads aligned to the position.  Comparison of the purity values between for all positions in the genome for two orthogonal sequencing methods was used to differentiate between positions with low purity values due to platform specific biases and those due to true biological diversity `r I(pasteLabel("Figure", figCount, "f.purity_scatter"))`.  Out of $`r rm_genome_size/1000000`\times 10^6$ positions in the genome `r pos_pur_gt99_both` positions had purity values greater than 99\% for both short read sequencing platforms `r I(pasteLabel("Table", tableCount, "t.purity_counts"))`.  `r pos_pur_gt99_one` and `r pos_pur_gt95_one` positions with purity values greater than 99% and 95% for one of the two platforms.  All postions had a purity value greater than `r min_max_pur_position` for at least one of the two platforms. Positions with the low purity values for one platform and not another indicate a potential platform specific bias.  The positions with low purity for MiSeq were non-uniformly distributed where as positions with low purity for PGM were unifromly distributed `r I(pasteLabel("Figure", figCount, "f.purity_genome_coord"))`.  The reference base is identified using a third orthogonal sequencing method (PacBio RSII) and thus a low value for one of the two short reads sequencing platforms represents agreement between two of the three sequencing methods.  

```{r cache = TRUE, echo = FALSE, message = FALSE, warnings = FALSE}
source("base_purity_analysis.R")
```

```{r echo = FALSE, message = FALSE, warnings = FALSE}
tableCount <- incCount(tableCount, "t.purity_counts")
pur_dat_id %>% group_by(pur_group) %>% summarise(count = n()) %>% kable(row.names = NA, caption = "Number of genome positions with purity values higher and lower than 0.99 for MiSeq and PGM sequencing platforms.")
```

```{r cache = TRUE, echo = FALSE, message = FALSE, warnings = FALSE, fig.show = 'asis', fig.cap= "_Comparison of base purity values for PGM and MiSeq._ Positions are colored based of high and low purity values for the two sequencing platforms."}
figCounts <- incCount(figCount, "f.purity_scatter")
ggplot2::ggplot(pur_dat_id_filt) +
    ggplot2::geom_point(ggplot2::aes(x = plat1, y = plat2, color = pur_group),
                        alpha = 0.5) +
    ggplot2::labs(x = "MiSeq", y = "PGM") + ggplot2::theme_bw() + 
    ggplot2::xlim(min_val, 1) + ggplot2::ylim(min_val, 1)
```

```{r cache = TRUE, echo = FALSE, message = FALSE, warnings = FALSE, fig.show = 'asis', fig.cap= "_Position of bases with low purity for one or both of the two short read sequencing platforms._"}
figCounts <- incCount(figCount, "f.purity_genome_coord")
ggplot(pur_dat_id_filt %>% filter(CHROM == chrom_names[1])) + 
    geom_bar(aes(x = POS, fill = pur_group)) + 
    facet_wrap(~pur_group, ncol = 1, scales = "free_y") + theme_bw()
```


### Sequence homogeneity
The genomic material homogeneity was assessed through a pairwise statistical analysis of the replicate MiSeq datasets using the varsan somatic variant caller.  The pairwise variant analysis failed to identify any statistically significant base level differences among the replicates `r I(pasteLabel("Table", tableCount, "t.homogeneity_sig"))`.

```{r echo =FALSE}
tableCount <- incCount(tableCount, "t.homogeneity_sig")
kable(homogeneity_sig_table(peprDB, rename_cols = TRUE), 
      row.names = FALSE,
      caption = "Pairwise variant analysis results")
```

## Genomic Purity
Based on taxonomic assignment of the PGM and MiSeq data the reference material has minimal if any genomic contaminants `r I(pasteLabel("Figure", figCount, "f.contam_counts"))`, with a maximum `r max_contam` of reads in any dataset classified as beloning not beloning to the genus `r rm_genus`.  The most abundant contaminant was _Escherichia coli_ which is a well documented contaminant of molecular biology reagents `r I(pasteLabel("Figure", figCount, "f.contam_taxa"))`. Lower abundance contaminants are most likely bioinformatic errors and not true contaminants. To provide a conservative estimate of the genomic purity the stated value assumes all reads identified as not beloning to the genus _Staphylococcus_ are contamiants.

```{r echo = FALSE, fig.show = 'asis', fig.cap= "Proportion of reads from contaminant DNA.", warning=FALSE, message=FALSE}
figCounts <- incCount(figCount, "f.contam_counts")
contam_counts_figure(peprDB, rm_genus)
```

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.height = 6, fig.cap= "Breakdown of contaminants by organism.",fig.retina=TRUE}
figCounts <- incCount(figCount, "f.contam_taxa")
contam_point_line_figure(peprDB, rm_genus)
```

# Discussion
## General Pipeline information
* Run time and system requirements
    * Pipeline is not optimized for runtime or storage
    * plans to add parallelization and restart functionality using ruffus

## Sequencing
The PEPR database includes a number of summary statistics for the sequencing datasets processed.  Sequencing dataset throughput and read length were as expected based on the library prep and sequencing methods used excluding the PGM datasets.  The 400 bp sequencing and library prepartion methods yielded median read lengths of `r pgm_med_read_length`. The shorter read length is potentially due to the low GC content is most likely due to low GC content is known to challenge current sequencing technologies sequencing [@Quail2012].  
__TODO Summary statement about min coverage TODO__

## Genome assembly
A closed reference genome assembly is required as input for the PEPR pipeline.  A closed candidate reference genome was assembled using PacBio long read data.  To assess accuracy of the reference genome whole genome mapping data generated with the OpGen optical mapping technology, an orthogonal measurement method, was used to evaluate the genome assembly. Optical mapping results were used to assess the overall structure of the genome, its use of long DNA fragments (__TODO fragment size TODO__) allow for the evaluation of large misassemblies , > 3 kb, that are not easily identified using standard short read sequencing data.  Mate pairs, or other library prepartion methods developed such as bionanogenomics (??), represent orthogonal methods that can also be used to identify large misassemblies.  The reference was then used as input for the PEPR evaluation pipeline.  

The evaluation step in the PEPR pipeline provides an automated method for evaluate and refine a reference genome.The evaluation pipeline takes a reference genome and uses the genome assembly polishing and assessment tool Pilon [@Walker2014].  Pilon was choosen for the genome evalaution step in the pipeline as it not only assesses the accuracy of the pipeline but also attempts to automatically correct errors in the assembly.  Other methods are available for evaluating reference genomes, e.g. amosvalidate [@Phillippy2008] and ALE [@Clark2013], however these methods only assess assembly accuracy and do not correct missassemblies.  To simplify the pipeline Pilon used as it evlauates accuracy and attempts to correct ambiguities. The agreement between the PacBio assembly, Optical Mapping data, and MiSeq Pilon analysis allows for increased confidence in the PacBio reference assembly.  The resulting reference assembly represents a consensus genome for the population of cells used to generate the material being characterized.  The pipeline does not attempt to identify or characterize low frequency structural variants within the material. Additionally, the pipeline was not developed to assess the homogeneity, vial-to-vial variability of the reference genome.  

## Base level analysis
The PEPR evaluation pipeline validated reference genome is then used as input for the characterization pipeline which calculates base level statistics using replicate sequencing data from orthogonal measurement methods. To assess the purity of a genome position or base, sequencing data from the replicate datasets are pooled by sequecing platform.  Samtools mpileup is used to calculate the number of base calls at that support or disagree with the reference base.  The base call counts are used to calculate the base level purity.  Through comparison of the base purity for two orthogonal sequencing methods we are able to identify genome positions low bases with low purity values due to platform specific systematic sequencing errors.  It is also important to acknowledge that the two platforms in agreement are susceptible to the same bias and thus the outlier platform could indicate the true value.  The sequencing technology used to chracterize the material is still maturing and an incomplete understanding of platform specific biases limits our ability to provide a confidence value for the base calls.

A number of base levels metrics, such as strain bias, are calculated as part of the PEPR pipeline and are included in the pipeline results database, and can be used to differentiate positions with low purity due to measurement error and thoes due to biological variability. Use of additional metrics and algorithms developed for the identification of low frequency variants, such as lowfreq.  Can help identify positions with low levels of biological variability.

The PEPR characterization pipeline assess the homogeneity of a material through comparison of purity values between technical replicates.  The homogeneity analysis is performed using the varscan tumor-normal variant caller [@Koboldt2009].  Only MiSeq data was used to assess the homogeneity of the material as the higher coverage increased the statistical power of the test and the replicate libraries provide information regarding the method error rate.  No statistically significant variants were identified between all pair-wise comparisons indicating that the material is homogeneous.  __TODO text multiple tests p-value corrections TODO__

## Genomic Purity
The third component of PEPR is the `genomic_purity` purity pipeline.  This component of the pipeline is used to identify DNA within the material that belongs to a genus other than the material genus.  Genomic contamiants can be from the culture itself or reagents and materials used to prepare the material [@Shrestha2013; @Tang2003; @Salter2014].  Contaminants identified by the `genomic_purity` pipeline may not necessarily be present in the material for example reagents used to during library preparation may also include contaminants [@Tanner1998; @Newsome2004; @Motley2014; @Salter2014].  Bioinformatic errors may also lead to false positive contaminants, these are either due to errors in the database, due to sequence misclassification, or errors in the classification algorithm itself.  Genomic purity analysis of the _S. aureus_ material identified a number of canidate contaminants,  the most abundant of which was _E. coli_.  _E. coli_ a well documented contaminant of molecular biology reagents, and not likely a true contaminant [@Salter2014]. While, contaminants identified by the `genomic_purity` are most likely from reagents and due to bioinformatic errors, a conservative value the material genomic purity value assumes all contaminants are real.


# Conclusions
PEPR provides an initial framework for chracterizing microbial genomic reference materials. The objective is developing the PEPR was to provide a reproducible and transparent workflow for the characterization of prokaryotic genomic material.  The pipeline can be used to characterize reference materials as well as in-house quality control materials for which sequencing data from multiple platforms is available for replicates sample.  The resulting characterization values are intentially conservative and without uncertainty or confidence estimates as sources of bias and error associated with the meausrement process are currently not fully understood. As the scientific communities understanding of the measurement process matures new algorithms can be incorporated into the pipeline increasing the quality of material characterization results. 

# Acknowlegements
The author’s would like to thank Jenny McDaniel, Lindsay Vang, and David Catoe for performing the measurements. Tim Muruvanda for performing the PacBio sequencing. This work was supported by the Department of Homeland Security (DHS) Science and Technology Directorate under the Interagency Agreement HSHQPM-14-X-00078 with NIST and by two interagency agreements with the FDA. Opinions expressed in this paper are the authors’ and do not necessarily reflect the policies and views of the DHS, NIST, or affiliated venues. Certain commercial equipment, instruments, or materials are identified in this paper only to specify the experimental procedure adequately. Such identification is not intended to imply recommendation or endorsement by the NIST, nor is it intended to imply that the materials or equipment identified are necessarily the best available for the purpose.
Official contribution of NIST; not subject to copyrights in USA.

# References

[^opgen]: OpGen Inc. [opgen.com](http://opgen.com) 708 Quince Orchard Road Gaithersburg, MD 20878 USA
[^illumina]: Illumina Inc., [illumina.com/](http://www.illumina.com/) 5200 Illumina Way San Diego, CA 92122 USA
[^iontorrent]: Life Technologies Corp., [iontorrent.com/](http://www.iontorrent.com/) 7000 Shoreline Court #201, South San Francisco, CA 94080 USA
[^pacbio]: Pacific Biosciences of California Inc. [pacificbiosciences.com/](http://www.pacificbiosciences.com/) 1380 Willow Rd. Menlo Park, CA 94025 USA